error checking the LE decomp functions

### 2016-09-22 Code/Test life expectancy fxs on Alabama.Rmd
improved LE functions and split files for testing

---
title: "error checking the LE decomp functions"
author: "Corinne Riddell"
date: "September 22, 2016"
output: html_document
---

**An example**

This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.

First, I create the dataset in R:
```{r create_dataset}
canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
                         "obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434, 3681, 5155, 8449, 13421, 19371, 
                                          25448, 31465, 37441, 48796, 64763, 72512, 60109, 43812),
                         "obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
                                        4532151, 5085092, 5163532, 4597916, 3937984, 3120981, 2323629, 1862971,
                                        1499907, 997492, 492445, 196109))

quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
                         "obs.deaths" = c(1057, 164, 116, 176, 659, 956, 1056, 1110, 1498, 2470, 4116, 6054, 8426,
                                          11175, 13558, 16567, 21349, 22396, 16851,10878),
                         "obs.popn" = c(215200, 789710, 1008266, 1179423, 1249360, 1261780, 1349857, 1293521,
                                        1313834, 1551684, 1622835, 1485249, 1288014, 1077253, 796557, 609341, 
                                        476211, 300506, 134353, 49678))
```

I then add a column that denotes how many years are in each age bracket:
```{r add_n_x}
canada.men$n_x <- 5
canada.men$n_x[canada.men$ages == 0] <- 1
canada.men$n_x[canada.men$ages == 1] <- 4

quebec.men$n_x <- 5
quebec.men$n_x[quebec.men$ages == 0] <- 1
quebec.men$n_x[quebec.men$ages == 1] <- 4

head(canada.men)
head(quebec.men)

cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")

```

This source file contains the life expectancy functions used in this project:
```{r load_functions}
source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
```

```{r calculate_life_expectancy}
life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")

round(life.table.cm, 2)

life.table.qm <- life.table(data = quebec.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")

round(life.table.qm, 2)
```

```{r calc_age_decomp}
age_decomp <- le_age_decomp(life.table.qm, life.table.cm)
age_decomp
#check:
sum(age_decomp$C_x)
age_decomp$Life_Expectancy_2[age_decomp$Ages ==0] - age_decomp$Life_Expectancy_1[age_decomp$Ages ==0]

#also compare to the decomposition if the data is entered in the reverse order:
age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm)
age_decomp2
#check:
sum(age_decomp2$C_x)
age_decomp2$Life_Expectancy_2[age_decomp2$Ages ==0] - age_decomp2$Life_Expectancy_1[age_decomp2$Ages ==0]

```

```{r calc_cod_decomp}
canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
                      cod.table = cod.props,age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
                      prop1.colname.cod.table = "prop1", prop2.colname.cod.table = "prop2")

cod.contribution$Cause.of.death <- factor(cod.contribution$Cause.of.death, 
                                          levels = cod.contribution$Cause.of.death[order(cod.contribution$cod.contribution,
                                                                                         decreasing = T)])

ggplot(cod.contribution, aes(x = Cause.of.death, y = cod.contribution)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_abline(intercept = 0, slope = 0)
```

**A second example**

In this example we use the data that Sam entered into Stata to illustrate the life expectancy concept to me. I'll replicate the life table here to ensure that I can arrive at the same results as he does in Stata.

```{r lifetable_cali_complete}
cali.complete <- data.frame("ages" = c(0:15), "nx" = rep(1,16),
                         "obs.deaths" = c(1635, 64, 41, 22, 41, 55, 42, 58, 44, 52, 48, 60, 52, 82, 129, 23300), #last number is fictitious
                         "obs.popn" = c(123342, 111520, 109200, 108749, 105698, 110548, 106857, 112184, 116423,
                                        132952, 134266, 128938, 125502, 128212, 132775, 143600))

head(cali.complete)

life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "obs.deaths", population.counts = "obs.popn")

cali.complete$avelived <- 0.5
cali.complete$avelived[cali.complete$ages==0] <- 0.1

life.table(data = cali.complete, age.groups = "ages", num.ages.in.group = "nx", ave.prop.lived = "avelived",
           death.counts = "obs.deaths", population.counts = "obs.popn")
```

```{r lifetable_cali_abridged, warning=F, message=F}
cali.complete$group <- cut(cali.complete$ages, breaks=c(0, 1, 5, 10, 15, 20), 
                           labels = c("<1 years old", "1-4 years old", "5-9 years old", 
                                      "10-14 years old", "15 or older"), include.lowest = T, right = F)

table(cali.complete$group, cali.complete$ages)

library(dplyr)
cali.abridged <- cali.complete %>% group_by(group) %>% 
                                   summarise(nx = sum(nx), ages = first(group), popn = sum(obs.popn), deaths = sum(obs.deaths),
                                             avelived = first(avelived))

cali.abridged

data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn"))

data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived"))
```

```{r alabama_example}
#load("Data/alabama_only.Rdata")
load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

@@

ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
  geom_point(aes(y = Life_Expectancy_2, col="White"))
```

>>>>

---
title: "Test functions on Alabama data"
author: "Corinne Riddell"
date: "September 22, 2016"
output: html_document
---

```{r alabama_example}
load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

@@

ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
  geom_point(aes(y = Life_Expectancy_2, col="White"))

alabama_cod_decomp <- cause_of_death_decomp(life.table1 = life.table.blacks, life.table2 = life.table.whites, decomp.table = alabama_age_decomp,
                      cod.table = cod.tab.alabama.male.1969, "Age2", "COD2", "prop1", "prop2")
```

### 2016-09-22 Code/error checking the LE decomp fxs.Rmd
improved LE functions and split files for testing

**An example**

This data was used in a previously published paper [@auger2014mortality]. We use it to show how the life expectancy function works.

First, I create the dataset in R:
```{r create_dataset}
canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
                         "obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434, 3681, 5155, 8449, 13421, 19371, 
                                          25448, 31465, 37441, 48796, 64763, 72512, 60109, 43812),
                         "obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
                                        4532151, 5085092, 5163532, 4597916, 3937984, 3120981, 2323629, 1862971,
                                        1499907, 997492, 492445, 196109))

@@

head(canada.men)
head(quebec.men)

cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")

```

This source file contains the life expectancy functions used in this project:
```{r load_functions}
source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
```

```{r calculate_life_expectancy}
life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", death.counts = "obs.deaths", population.counts = "obs.popn")

@@

```{r calc_age_decomp}
age_decomp <- le_age_decomp(life.table.qm, life.table.cm)
age_decomp
#check:
sum(age_decomp$C_x)
age_decomp$Life_Expectancy_2[age_decomp$Ages ==0] - age_decomp$Life_Expectancy_1[age_decomp$Ages ==0]

#also compare to the decomposition if the data is entered in the reverse order:
age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm)
age_decomp2
#check:
sum(age_decomp2$C_x)
age_decomp2$Life_Expectancy_2[age_decomp2$Ages ==0] - age_decomp2$Life_Expectancy_1[age_decomp2$Ages ==0]

```

```{r calc_cod_decomp}
canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
                      cod.table = cod.props,age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
                      prop1.colname.cod.table = "prop1", prop2.colname.cod.table = "prop2")

cod.contribution$Cause.of.death <- factor(cod.contribution$Cause.of.death, 
                                          levels = cod.contribution$Cause.of.death[order(cod.contribution$cod.contribution,
                                                                                         decreasing = T)])

@@

```{r alabama_example}
#load("Data/alabama_only.Rdata")
load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]

second.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.Black.Male", ]

alabama.by.strata.cod <- dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
                                               arrange(RaceSex, Year2, Age2) %>%
                                               mutate(total_deaths5 = sum(Count_md5),
                                                      cod_prop_death5 = Count_md5/total_deaths5)

first.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.White.Male", ]
second.strata.cod <- alabama.by.strata.cod[alabama.by.strata.cod$StateYearRaceSex == "Alabama.1969.Black.Male", ]

cod.tab.alabama.male.1969 <- first.strata.cod %>% ungroup() %>%
                                                  select(Age2, COD2, cod_prop_death5) %>%
                                                  rename(prop2 = cod_prop_death5)

cod.tab.alabama.male.1969$prop1 <- second.strata.cod$cod_prop_death5
cod.tab.alabama.male.1969 <- cod.tab.alabama.male.1969 %>% arrange(COD2, Age2)

ggplot(cod.tab.alabama.male.1969, aes(x=COD2, y=prop1)) + 
  geom_point(aes(col="White")) + geom_point(aes(y=prop2, col="Black")) +
  scale_y_continuous(name = "Proportion dying from specified cause") +
  facet_wrap(~Age2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

system.time(life.table.whites <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))

life.table.blacks <- data.frame(life.table(data = second.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))

ggplot(life.table.blacks, aes(x = Age2, y = q_x)) + geom_point() + geom_point(aes(y = R_x), col="red")

life.table.whites[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]

system.time(
for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
  assign(paste0("lt.", as.character(stratum.id)), 
         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
                                  age.groups = "Age2", num.ages.in.group = "nx", 
                                  death.counts = "total_deaths5", population.counts = "Population"))
}
)

alabama_age_decomp <- le_age_decomp(life.table.blacks, life.table.whites, "Age2")
alabama_age_decomp

ggplot(alabama_age_decomp, aes(y = C_x, x = Ages)) + geom_point() + scale_y_continuous("Contribution of Age to Gap")
sum(alabama_age_decomp$C_x)
life.table.whites$e_x[1] - life.table.blacks$e_x[1]

ggplot(alabama_age_decomp, aes(y = Life_Expectancy_1, x = Ages)) + geom_point(aes(col="Black")) + scale_y_continuous("Life Expectancy (Yrs)") +
  geom_point(aes(y = Life_Expectancy_2, col="White"))
```

>>>>

In this file we apply the functions created to calculate and decompose life expectancy by way of example, and as a way to cross-check against results from published findings (to ensure my functions work properly).

Our functions:
```{r source_LE_fxs}
library(dplyr)
library(ggplot2)
source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
```

**First example**

We use data from a paper published by Auger et al and compare our findings to those found by the authors.

Here is the data need to calculate life expectancy, as provided by the authors in a supplementary appendix (xls file) to their paper:
```{r create_dataset}
canada.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
                         "obs.deaths" = c(4054, 625, 445, 622, 2568, 3770, 3434, 3681, 5155, 8449, 13421, 19371, 
                                          25448, 31465, 37441, 48796, 64763, 72512, 60109, 43812),
                         "obs.popn" = c(719953, 2835604, 3654546, 4071400, 4464549, 4564679, 4296613, 4237402,
                                        4532151, 5085092, 5163532, 4597916, 3937984, 3120981, 2323629, 1862971,
                                        1499907, 997492, 492445, 196109))

@@

head(canada.men)
head(quebec.men)
```

We also need data on the proportion of deaths within each age bracket due to each cause of death.
```{r data_cod_proportions}
cod.props <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteGap/working_teaching_examples/auger_cod_proportions.csv")
names(cod.props)[3] <- "prop.QC"
names(cod.props)[4] <- "prop.CAD"
head(cod.props)
```

Calculate and display the life tables for the Canada and Quebec data:
```{r calculate_life_expectancy}
life.table.cm <- life.table(data = canada.men, age.groups = "ages", num.ages.in.group = "n_x", 
                            death.counts = "obs.deaths", population.counts = "obs.popn")

@@

Life expectancy at birth in Canadian men is `r round(life.table.cm$e_x[1], 2)` vs. `r round(life.table.qm$e_x[1], 2)` in Quebec men, for a life expectancy gap of `r life.table.cm$e_x[1] - life.table.qm$e_x[1]` years.

Decompose the life expectancy gap by age:

```{r calc_age_decomp}
age_decomp <- le_age_decomp(life.table1 = life.table.qm, name.lt1 = "Quebec", 
                            life.table.cm, name.lt2 = "Canada", 
                            age.groups = "ages")
age_decomp
sum(age_decomp$C_x) 
```

Check that the function still works when the life tables are provided in the reverse order:

```{r calc_age_decomp_reverse}
age_decomp2 <- le_age_decomp(life.table1 = life.table.cm, name.lt1 = "Canada", 
                             life.table2 = life.table.qm, name.lt2 = "Quebec", 
                             age.groups = "ages")
age_decomp2
sum(age_decomp2$C_x)
```

Compare the output of the above two functions -- the contribution of age to the LE gap are slightly different based on the order of operation:

```{r compare decomp_output}
round(cbind(age_decomp$C_x, age_decomp2$C_x),4)
```

```{r calc_cod_decomp}
canada.cod.decomp <- cause_of_death_decomp(life.table1 = life.table.qm, life.table2 = life.table.cm, decomp.table = age_decomp,
                      cod.table = cod.props, age.colname.cod.table = "Age", COD.colname.cod.table = "cod", 
                      prop1.colname.cod.table = "prop.QC", prop2.colname.cod.table = "prop.CAD")
```

Collapse by cause of death to calculate the overall contribution of each cause of death to the gap in life expectnancy:
```{r }
cod.contribution <- canada.cod.decomp %>% group_by(Cause.of.death) %>% summarise(cod.contribution = sum(C_xi))

#order by contribution amount for graphing purposes
cod.contribution$Cause.of.death <- factor(cod.contribution$Cause.of.death, 
                                          levels = cod.contribution$Cause.of.death[order(cod.contribution$cod.contribution,
                                                                                         decreasing = T)])

@@
